---
title: "sarima"
format: pdf
---

Libraries
```{r}
library(here)
library(astsa)
library(tidyverse)
library(forecast)
```

Loading data
```{r}
df <- arrow::read_parquet(here("Monthly_Modal_Time_Series.parquet"))
head(df)
```
Filtering for Pittsburgh Regional Transit, bus system ('Motor Bus'), 2021-2025
```{r}
burgh.df <- df[df$Agency == 'Pittsburgh Regional Transit', ]
burgh.bus.df <- burgh.df[burgh.df$Mode == 'MB', ]
burgh.bus.df <- burgh.bus.df[burgh.bus.df$Year >= 2021 & burgh.bus.df$Year <= 2025, ]
final.df <- burgh.bus.df[, c('Year', 'Month', 'Unlinked Passenger Trips'), ]
rownames(final.df) <- NULL
ts.df <- ts(final.df$'Unlinked Passenger Trips',
            start = c(2021, 1),
            end = c(2025, 12),
            frequency = 12)
head(ts.df)
```
## Model 1: ARIMA, faster search, no seasonal component
```{r}
model1 <- auto.arima(ts.df, 
                  stepwise = TRUE, 
                  approximation = TRUE,
                  seasonal = FALSE,
                  ic = "aic")
summary(model1)
```
This algorithm suggests an ARIMA(0,1,3) model.
```{r}
resi_model1 <- ts.df - model1$fitted
tsdisplay(resi_model1)
```
These residuals are clearly not white noise (which makes sense, since we can assume intuitively that there will be a seasonal component here)
```{r}
plot(model1$fitted,col = "red")
lines(ts.df,col = "black")
```
These predictions don't look horrific, but they seem a bit "delayed".
## Model 2: SARIMA, faster search
```{r}
model2 <- auto.arima(ts.df, 
                  stepwise = TRUE, 
                  approximation = TRUE,
                  seasonal = TRUE,
                  ic = "aic")
summary(model2)
```
This algorithm suggests a $SARIMA(0,1,0)(1,1,0)_{12}$ model.
```{r}
resi_model2 <- ts.df - model2$fitted
tsdisplay(resi_model2)
```
These residuals look more like white noise!
```{r}
plot(model2$fitted,col = "blue")
lines(ts.df,col = "black")
```
These predictions look better than the previous model, with the predicted spikes coming at the right times instead of consistently after a spike is observed.

## Model 3: SARIMA, exhaustive search
```{r}
model3 <- auto.arima(ts.df, 
                  stepwise = FALSE, 
                  approximation = FALSE,
                  seasonal = TRUE,
                  ic = "aic")
summary(model3)
```
This algorithm suggests a $SARIMA(0,1,0)(0,1,1)_{12}$ model.
```{r}
resi_model3 <- ts.df - model3$fitted
tsdisplay(resi_model3)
```
Residuals look pretty good again.
```{r}
plot(model2$fitted,col = "darkorange")
lines(ts.df,col = "black")
```
This doesn't seem that different from Model 2 to me.


